---
title: "Ggplot part 2"
author: "Alys Young and the coding committee"
date: "`r Sys.Date()`"
output: html_document
---

Alys to look into:
Packages: "ggforce", "gghighlight", "ggnewscale", "ggplot2", "ggraph", "ggrepel", "ggtext", "ggthemes",

Today's session:  

- other packages building on ggplot
- more model plots
- colour palettes
- add ins
- custom themes

Haven't covered spatial data (e.g maps) - this will come in a future session !



## Modelling
Today we are going to keep looking at the linear model which predicts the sepal width with variables within the iris package.
```{r model_setup}
library(ggplot2)
library(dplyr)
data_model <- iris %>% 
  mutate(petal_shape = Petal.Length / Petal.Width)

m <- lm(Sepal.Width ~ Sepal.Length + Petal.Length + Petal.Width + Species + petal_shape + Sepal.Length:Species, data = data_model)
```


The ggeffects package can be used to plot the marginal effects of a variable of interest. Note, these plots Mean values of the other variables.
```{r effect}
library(ggeffects)

ggpredict(m, terms = "Petal.Length" )

marginalEffect_df <- ggpredict(m, terms = "Petal.Length" )

plot(marginalEffect_df)

ggplot(data = marginalEffect_df, aes( x = x, y = predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin =  conf.low, ymax = conf.high), alpha = .1)
```


```{r effect2}

marginalEffect_df <- ggpredict(m, terms = c("Petal.Length", "Species", "Sepal.Length"))

plot(marginalEffect_df)

ggplot(data = marginalEffect_df, aes( x = x, y = predicted, fill = group, colour = group)) +
  geom_line() +
  geom_ribbon(aes(ymin =  conf.low, ymax = conf.high), alpha = .1) +
  facet_wrap(~facet) +
  labs(x = "Petal length", y = "Predicted sepal width", colour = "Species", fill = "Species", title = "Predicted sepal width by petal length, species and sepal length")
```



Correlation plot

```{r corrlation}

# select only the numeric data to test for correlations
# Option 1 in base R
  # using an apply function, test each column to see if its numeric, and then select only those
data_num <- data_model[sapply(data_model, is.numeric)]

# option 2 using dplyr
data_num <- data_model %>% dplyr::select_if(is.numeric)

## Correlation
cor_out <- cor(data_num)

## Plot the correlation
ggcorrplot(cor_out)
ggcorrplot(cor_out,  
           type = "lower",
           lab = TRUE)

```


### Choosing your own colours
Single colours
Colour blind friendly
- https://cran.r-project.org/web/packages/colorBlindness/vignettes/colorBlindness.html
- packages such as: minsell, viridis, RColorBrewer, dichromat, colourblindr, shades, ggsci
Premade colour palettes
Making your own colour palette

```{r colourblind_check}

library(colorBlindness)

p <- ggcorrplot(cor_out)
cvdPlot(p)

```


```{r}
# Colour by a name and shape
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(shape = 17, color  = "darkgreen")

# Colour by a HEX and size
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(alpha = 0.5, color = "#4842f5", size = 0.8)
```



Load premade palettes
```{r cols}
# install.packages("RColorBrewer")
# install.packages("viridisLite")
library(RColorBrewer)
library(viridisLite)

# To see the color palettes from RColorBrewer
display.brewer.all()
```

When you need to "fill" for a categorical variable:
```{r cols_fill_cat}
# categorical variables
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()

# Colour by name or HEX
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_manual(values = c("darkgreen", "purple", "#4842f5"))

# Colour from a preloaded palette
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_brewer(palette="Set2")

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_viridis_d() # d for discrete

# Grey
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_grey()

```


When you need to "fill" for a continuous variable:
```{r cols_fill_cont}
# categorical variables
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point()

# Colour by name or HEX
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_gradient(low="blue", high="red")

# Colour from a preloaded palette
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_gradientn(colours = rainbow(5))

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_distiller(palette = "Greens") # the palette is chosen from display.brewer.all()

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_viridis_c() # c for continuous

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_viridis_c(option = "plasma") # choosing a palette. try "magma", or "inferno"

```

When you need to "color":
```{r cols_color}
# categorical variables
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot()

# Colour by name or HEX
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_manual(values = c("green", "pink", "#4842f5"))

# Colour from a preloaded palette
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_brewer(palette="Set2")

ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_viridis_d()

# Grey
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_grey()

```

## Add ins
ggExtra

ggthemeassist

ggthemes

### Themes and customising your graph
Text size and angled. At the top of the script, using `theme_set()` with your specifications.
```{r text modify}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  theme_classic() +
  theme(text = element_text(size = 15), axis.text.x = element_text(angle = 45, hjust = 1) ) 
  
```


### Integration with other packages
### Plotly
 javascript visualisation toolkit with an R interface. It’s a great tool if you want to make interactive graphics for HTML documents
 
```{r ggplotly}
# install.packages("plotly")
library(plotly)

p <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

ggplotly(p)
  
```
Change the text for the mouse hover
```{r ggplotly2}
p <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, text = paste0("Sepal length: ", Sepal.Length, "cm <br>", "Sepal width: ", Sepal.Width, "cm <br>", "Petal length: ", Petal.Length, "cm <br>", "Petal width: ", Petal.Width, "cm <br>", "Species: ", Species) )) +
  geom_point()

ggplotly(p, tooltip = "text")
  
```