---
title: "Intro to Ggplot"
author: "Simone and Alys"
date: "28/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ggplot

When you begin to need more functionality than Base R plotting can give you, its time to turn to ggplot. Infintely more tweakable and incredibly well documented, with ggplot you can create sleak and paper-quality plots and graphs. 

This tutorial is an introduction to ggplot with examples to get you started.

### Setting up

```{r load, message=TRUE, warning=FALSE}
# install.packages("ggplot2")
library(ggplot2)
```

## Plots 101

You need to:
1. Start by calling a ggplot using `ggplot()`
2. Specify the data using the "data" argument `ggplot(data =  )`, and the x and y variables using the aesthetics argument   `ggplot(data = , aes(x= , y = ) )`
3. Specify  what sort of plot you make ggplot to make using a second line joing to the `ggplot()` with a `+`

Lets show some examples using the cars dataset pre-loaded in R

```{r data}
# load the data
data(iris) 

# Take a look at the data
head(iris)
```
For  all the ggplots with this data, the first line will look like this, with the x and y variables changing depending on what we want to plot:
```{r set-up}
ggplot(data = iris, aes(x = , y = ))
```

### Basic sorts of plots

One categorical variable and one continuous
relationship between Sepal.Width by the species
```{r cat}
# Box Plot 
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot()

# Voilin plot
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_violin()

# Bar plot
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_bar(stat = "identity")

# a dot plot
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()
```




### Add another variable
You can display another variable on the graph too. Commonly use the colour, also can you shape or size

But pay attention to the difference between color and fill. Also this is different for the type of graph 
```{r vars}

# A bubble plot with size for a continuous variable
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,  size = Petal.Length)) +
  geom_point(alpha = 0.5)

# A bubble plot with size for a categorical variable
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,  size = Species)) +
  geom_point(alpha = 0.5)

# Change the shape for a categorical variable
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,  shape = Species)) +
  geom_point()

# Fill for violin on a categorical variable
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill  = Species)) +
  geom_violin()

# Color for violin  on a categorical variable
ggplot(data = iris, aes(x = Species, y = Sepal.Width, color  = Species)) +
  geom_violin()

# Color for point on a continuous variable
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,  color = Sepal.Length)) +
  geom_point()

# Color for point on a categorical variable
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,  color = Species)) +
  geom_point()


```

You can also use mutliple of these togeher, either for the same or different variables. However, be sure that adding more is actually enhancing the point you are making with the graph, and not making it more difficult to understand.


```{r vars2}

# categorical variables
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species, shape = Species)) +
  geom_point()

# continuous variables
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length, size = Petal.Length)) +
  geom_point(alpha = 0.5)

```
You can also set these parameters to be constant and without being associated to a variable by not putting them into the aes() call, and instead specifying what you want it to look like.

```{r}
# Colour by a name and shape
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(shape = 17, color  = "darkgreen")

# Colour by a HEX and size
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(alpha = 0.5, color = "#4842f5", size = 0.8)
```


### Choosing your own colours
Load premade palettes
```{r cols}
# install.packages("RColorBrewer")
# install.packages("viridisLite")
library(RColorBrewer)
library(viridisLite)

# To see the color palettes from RColorBrewer
display.brewer.all()
```

When you need to "fill" for a categorical variable:
```{r cols_fill_cat}
# categorical variables
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()

# Colour by name or HEX
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_manual(values = c("darkgreen", "purple", "#4842f5"))

# Colour from a preloaded palette
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_brewer(palette="Set2")

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_viridis_d() # d for discrete

# Grey
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_grey()

```


When you need to "fill" for a continuous variable:
```{r cols_fill_cont}
# categorical variables
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point()

# Colour by name or HEX
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_gradient(low="blue", high="red")

# Colour from a preloaded palette
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_gradientn(colours = rainbow(5))

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_distiller(palette = "Greens") # the palette is chosen from display.brewer.all()

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_viridis_c() # c for continuous

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() +
  scale_color_viridis_c(option = "plasma") # choosing a palette. try "magma", or "inferno"

```

When you need to "color":
```{r cols_color}
# categorical variables
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot()

# Colour by name or HEX
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_manual(values = c("green", "pink", "#4842f5"))

# Colour from a preloaded palette
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_brewer(palette="Set2")

ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_viridis_d()

# Grey
ggplot(data = iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_boxplot() +
  scale_fill_grey()

```



### Basic modifcations
Add a title
```{r title}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()+
  labs(title = "Sepal length by width")
```

Change the x and y labels
```{r labs}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()+
  labs(title = "Sepal length by width", x = "Sepal length (cm)", y="Sepal width (cm)")
```

Lengend position
```{r legend}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  theme(legend.position = "bottom") # try "top"
```


Add a regression line to see trends
```{r scatter}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  geom_smooth()

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_manual(values=c('red','#E69F00', '#56B4E9')) + 
  geom_smooth(method = lm)
```

### Customise further
add other arguments for customisation

Box plot with flipped coordinates
```{r box flip}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  coord_flip()
  
```

Change the theme to black and white
```{r box theme}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  theme_bw() # try theme_light, theme_dark, theme_classic,

?theme_bw()
  
```

Text size and angled
```{r text modify}
ggplot(data = iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot() +
  theme_classic() +
  theme(text = element_text(size = 15), axis.text.x = element_text(angle = 45, hjust = 1) ) 
  
```


Facetting into column
```{r facet1}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  theme_bw() +
  facet_wrap(~Species, ncol = 3) # try ncol = 1, or instead of ncol write nrow
  
```

Facetting into rows with your own labels
```{r facet2}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  theme_bw() +
  facet_wrap(~Species, ncol = 1,
             labeller = labeller(Species = c("setosa" = "Iris setosa",
                                             "versicolor" = "Iris versicolor",
                                             "virginica" = "Iris virginica")))
  
```


Breaks and grid lines
```{r breaks}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  scale_x_continuous(limits = c(4, 8),
                     breaks = seq(4, 8, 1), # sequence from 0 to 8 by 1. Equivalent to 0:8
                     minor_breaks = seq(0, 8, 0.2)
                     )
  
  
```

Annotate
```{r annotate}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  annotate("text", x = 4.5, y = 4, label = "Group 1", size = 7) +
  annotate("text", x = 6.5, y = 3.6, label = "Group 2", color = "blue") + 
  annotate("text", x = 5.5, y = 2, label = "Group 3") 
  
# Bar plot
ggplot(data = iris, aes(x = Species, y = Sepal.Length)) +
  geom_bar(stat = "identity") +
    annotate("text", x = 1, y = 270, label = "*", size = 20) 
  
```

Draw arrows
```{r draw_lines}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  annotate("segment", x = 4, xend = 4, y = 2.8, yend = 3.5)

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  annotate("text", x = 4.5, y = 4.2, label = "Important segment") +
  annotate("segment", x = 4.1, xend = 5, y = 4, yend = 4,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm")))

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  annotate("text", x = 7, y = 4.5, label = "Outliers", size = 7) + # try size = 3
  annotate("segment", x = 7.4, xend = 7.8, y = 4.5, yend = 3.9, # try chaning these values
           size = 1.5, # try 0.5 or 2
           color = "darkblue", # try "green" or "purple"
           arrow = arrow())
  
```

### Building  a plot up step by step
```{r building up}
p <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

p

p + theme_bw()
p2 <- p + theme_bw()

p2 + labs(title = "Add title", subtitle = "And a sub-title", y = "Sepal width (cm)", x = "Sepal length (cm)")
  
```


## Save your plot
```{r save}
# ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
#  geom_point()

# ggsave()

```


### Integration with other packages
### Building  a plot up step by step
```{r ggplotly}
# install.packages("plotly")
library(plotly)

p <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

ggplotly(p)
  
```
Change the text for the mouse hover
```{r ggplotly2}
p <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, text = paste0("Sepal length: ", Sepal.Length, "cm <br>", "Sepal width: ", Sepal.Width, "cm <br>", "Petal length: ", Petal.Length, "cm <br>", "Petal width: ", Petal.Width, "cm <br>", "Species: ", Species) )) +
  geom_point()

ggplotly(p, tooltip = "text")
  
```